---
title: "Green Space Inequality via Spatial Income Patterns."
author: "Xun Lei"
affiliation: "Georgetown University"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left       
    number-sections: true   

execute:
  echo: false       
  warning: false    
  message: false   
---


# Introduction

Urban green space—such as parks, gardens, and recreational areas—is widely recognized as a critical component of environmental quality and public well-being. Access to green space has been linked to improved physical and mental health, greater opportunities for recreation, lower levels of urban heat exposure, and enhanced social cohesion within communities. As cities grow and densify, the equitable distribution of parks becomes a central concern in urban planning and environmental justice.

Many studies in environmental justice have found that socially advantaged neighborhoods often enjoy better access to ecological and recreational amenities, while lower-income communities may face “green space deficits.” These disparities can contribute to unequal health outcomes, reduced quality of life, and structural inequities embedded within the urban landscape.

Washington, D.C. provides a compelling case study for examining such spatial inequalities. The city exhibits substantial variation in median household income across its census tracts, with pronounced socio-economic differences between Northwest D.C. and the neighborhoods east of the Anacostia River. At the same time, the District maintains a large inventory of parks and green spaces, making it an ideal environment for studying whether these amenities are evenly—or unevenly—distributed.

This project investigates the spatial distribution of parks in Washington, D.C. and evaluates whether higher-income neighborhoods receive disproportionately greater access to green space. Specifically, the study focuses on first-order spatial properties through the estimation of an inhomogeneous intensity function and employs simulation-based hypothesis testing to assess whether observed patterns differ significantly from those expected under a population-driven null model.

By integrating open-source geospatial data with point-pattern methods from spatial statistics, this research contributes to understanding the extent of environmental inequality in the nation’s capital and highlights broader implications for urban planning and equitable resource allocation.

---

# Literature Review

Studies in environmental justice consistently show that access to parks and green space is often uneven across socio-economic groups. Research has found that higher-income neighborhoods typically have more and higher-quality green amenities, while lower-income communities face green space deficits (Wolch et al. 2014). Similar patterns have been documented in several U.S. cities, including evidence that park access strongly correlates with neighborhood wealth and historic patterns of urban inequality (Boone et al. 2009).  

This project builds on that literature by examining whether Washington, D.C. exhibits similar income-based disparities in the spatial distribution of parks.

---

# Methodology

This study evaluates whether park access in Washington, D.C. exhibits income-based spatial inequality using a combination of exploratory spatial analysis, spatial autocorrelation statistics, and simulation-based hypothesis testing. The methodological workflow follows the first-order spatial analysis framework emphasized in GIS for Spatial Data Science.

---

## Data Sources and Preprocessing

Two primary datasets were used:

1. **Park locations** from OpenStreetMap (OSM), represented both as polygons and as centroid points.
2. **Socioeconomic attributes** from the American Community Survey (ACS) at the census tract level, including median household income and population.

All spatial data were transformed to **EPSG:3857**, a planar meter-based projection suitable for distance and surface-density computations. For each census tract, park areas were intersected and aggregated to compute:

- **Total park area per tract (sqm)**
- **Park density**: park area per capita  
$$
\text{park density} = \frac{\text{park area}}{\text{population}}
$$


These variables provide the socio-economic and environmental covariates necessary for subsequent analysis.

---

# Exploratory Data Analysis (EDA)

```{r, include=FALSE}
library(sf)
library(dplyr)
library(ggplot2)
library(viridis)
library(sp)
library(spdep)
library(MASS)
library(gridExtra)
library(tidyr)
# Load data
tracts <- st_read("data/tract_parks.geojson")
parks_centroid <- st_read("data/parks_centroid.geojson")

# CRS
tracts <- st_transform(tracts, 3857)
parks_centroid <- st_transform(parks_centroid, 3857)

# Convert numeric
tracts <- tracts %>%
  mutate(
    median_income       = as.numeric(median_income),
    population          = as.numeric(population),
    total_park_area_sqm = as.numeric(total_park_area_sqm),
    park_density        = total_park_area_sqm / population
  )
```


```{r, echo=FALSE}
tracts_sp <- as_Spatial(tracts)
nb <- poly2nb(tracts_sp)
lw <- nb2listw(nb, style = "W")

moran.test(tracts$median_income, lw)
moran.test(tracts$park_density, lw)
```


```{r income-quartiles, echo=FALSE, fig.cap="Median Income Quartiles", fig.align="center"}

# Compute quartiles
tracts$income_quartile <- cut(
  tracts$median_income,
  breaks = quantile(tracts$median_income, probs = seq(0, 1, 0.25), na.rm = TRUE),
  include.lowest = TRUE,
  labels = c("Q1: Lowest", "Q2", "Q3", "Q4: Highest")
)

# Plot quartiles
ggplot(tracts) +
  geom_sf(aes(fill = income_quartile), color = "white", size = 0.1) +
  scale_fill_brewer(palette = "YlOrBr", name = "Income Quartile") +
  ggtitle("Median Income Quartiles Across Washington, D.C.") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 15, face = "bold", hjust = 0.5),
    legend.position = "right"
  )
```

```{r park-counts, echo=FALSE, fig.cap="Park Counts per Census Tract", fig.align="center"}

# Spatial join: assign each park to a tract
parks_in_tract <- st_join(parks_centroid, tracts, join = st_within)

# Count how many parks fall into each tract
park_counts <- parks_in_tract %>%
  st_drop_geometry() %>%
  count(GEOID, name = "n_parks")

# Merge counts to tracts
tracts <- tracts %>%
  left_join(park_counts, by = "GEOID") %>%
  mutate(n_parks = ifelse(is.na(n_parks), 0, n_parks))

# Plot
ggplot(tracts) +
  geom_sf(aes(fill = n_parks), color = "white", size = 0.1) +
  scale_fill_viridis_c(name = "Number of Parks", option = "C") +
  ggtitle("Park Counts per Census Tract") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 15, face = "bold", hjust = 0.5)
  )



```



# Hypothesis Testing

### Data Loading and Projection

```{r}
library(sf)
library(dplyr)
library(MASS)
library(viridis)
library(ggplot2)

tracts <- st_read("data/tract_parks.geojson") %>%
  st_transform(3857) %>%
  mutate(median_income = as.numeric(median_income))

parks_centroid <- st_read("data/parks_centroid.geojson") %>%
  st_transform(3857)

dc_boundary <- st_union(tracts)

coords <- st_coordinates(parks_centroid)
n_parks <- nrow(parks_centroid)

x_range <- range(coords[,1])
y_range <- range(coords[,2])
```


---

### Observed KDE of Park Locations

```{r}
library(fields)

obs_kde <- kde2d(
  x = coords[,1],
  y = coords[,2],
  n = 200,
  lims = c(x_range[1], x_range[2],
           y_range[1], y_range[2])
)

image.plot(
  obs_kde,
  col = viridis(200),
  main = "Observed Park Intensity (KDE)"
)

points(coords, pch = 16, cex = 0.3, col = "red")

```

**Findings:**
The KDE surface reveals visually distinct clusters of parks, especially in central and northwestern Washington, D.C. This provides initial evidence that park locations exhibit spatial concentration rather than a uniform distribution.

---



### Linking KDE Grid With Census Attributes

```{r}
grid_pts <- st_as_sf(
  expand.grid(x = obs_kde$x, y = obs_kde$y),
  coords = c("x", "y"),
  crs = st_crs(tracts)
)

nearest_tracts <- st_nearest_feature(grid_pts, tracts)

grid_df <- expand.grid(
  x = obs_kde$x,
  y = obs_kde$y
) %>%
  as_tibble() %>%
  mutate(
    obs_intensity = as.vector(obs_kde$z),  
    nearest_income     = tracts$median_income[nearest_tracts],
    nearest_population = tracts$population[nearest_tracts]
  )

```

**Findings:**
This code assigns each spatial grid cell the income value of its nearest census tract. This enables formal statistical testing between park density (continuous spatial surface) and neighborhood socioeconomic characteristics.

---

### Correlation Test

```{r}
cor_test <- cor.test(grid_df$obs_intensity, grid_df$nearest_income)
cor_test
```

**Findings:**
The Pearson correlation coefficient is **0.142**, with a p-value < 2.2e−16. Although the effect size is small, the relationship is statistically significant, indicating that higher-income neighborhoods tend to correspond to slightly higher park intensity.

---

### High-Income vs Low-Income T-Test

```{r}
income_median <- median(grid_df$nearest_income, na.rm = TRUE)

high_income_grid <- grid_df %>%
  filter(nearest_income >= income_median)

low_income_grid <- grid_df %>%
  filter(nearest_income < income_median)

t_test <- t.test(
  high_income_grid$obs_intensity,
  low_income_grid$obs_intensity
)

t_test
```

**Findings:**
The Welch t-test shows a highly significant difference between the two groups (p < 2.2e−16). High-income areas have a notably higher mean KDE intensity, demonstrating that park density systematically varies across socioeconomic strata.

---


### Scatter Plot: Income vs Park Intensity

```{r}
### ============================================================
### 1. Data Cleaning
### ============================================================

library(dplyr)
library(ggplot2)

# 1. Convert population to numeric
grid_df$nearest_population <- as.numeric(grid_df$nearest_population)

# 2. Replace impossible negative income values with NA
grid_df <- grid_df %>%
  mutate(nearest_income = ifelse(nearest_income < 0, NA, nearest_income))

# 3. Drop NA rows
grid_df <- grid_df %>% 
  tidyr::drop_na(nearest_income, nearest_population)

### ============================================================
### 2. Log Transformation
### ============================================================

grid_df <- grid_df %>%
  mutate(
    log_intensity = log(obs_intensity + 1e-12)   # avoid log(0)
  )

### ============================================================
### 3. OLS Regression on Log Intensity
### ============================================================

model_log <- lm(log_intensity ~ nearest_income, data = grid_df)
summary(model_log)

### ============================================================
### 4. Plot: log-transformed regression
### ============================================================

p_log <- ggplot(grid_df, aes(x = nearest_income, y = log_intensity)) +
  geom_point(alpha = 0.2, size = 0.8) +
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(
    title = "Log(Park Intensity) vs Income (OLS)",
    x = "Median Income ($)",
    y = "log(Park Intensity)"
  ) +
  theme_minimal()

p_log


### ============================================================
### 5. Income Grouping (high vs low income)
### ============================================================

income_median <- median(grid_df$nearest_income, na.rm = TRUE)

grid_df <- grid_df %>% 
  mutate(
    income_group = ifelse(nearest_income >= income_median, 
                          "High income", "Low income")
  )

### ============================================================
### 6. Grouped regressions
### ============================================================

# High income model
model_high <- lm(log_intensity ~ nearest_income, 
                 data = grid_df %>% filter(income_group == "High income"))

# Low income model
model_low <- lm(log_intensity ~ nearest_income, 
                data = grid_df %>% filter(income_group == "Low income"))

summary(model_high)
summary(model_low)

### ============================================================
### 7. Plot: grouped regression lines
### ============================================================

p_group <- ggplot(grid_df, aes(x = nearest_income, 
                               y = log_intensity, 
                               color = income_group)) +
  geom_point(alpha = 0.15, size = 0.7) +
  geom_smooth(method = "lm", se = FALSE, size = 1.2) +
  scale_color_manual(values = c("High income" = "red", "Low income" = "blue")) +
  labs(
    title = "Income-Stratified Regression: Log(Park Intensity)",
    x = "Median Income ($)",
    y = "log(Park Intensity)",
    color = "Income Group"
  ) +
  theme_minimal()

p_group


### ============================================================
### 8. Optional: Multiple Regression (income + population)
### ============================================================

model2 <- lm(log_intensity ~ nearest_income + nearest_population, 
             data = grid_df)
summary(model2)



```


---

### **Conclusion**


---

# Discussion



---

# Conclusion



---

# References

Wolch, Jennifer R., Jason Byrne, and Joshua P. Newell. 2014. “Urban Green Space, Public Health, and Environmental Justice: The Challenge of Making Cities ‘Just Green Enough.’” Landscape and Urban Planning 125: 234–244.

Boone, Christopher G., Geoffrey L. Buckley, J. Morgan Grove, and Chona Sister. 2009. “Parks and People: An Environmental Justice Inquiry in Baltimore, Maryland.” Annals of the Association of American Geographers 99(4): 767–787.

Rigolon, Alessandro. 2016. “A Complex Landscape of Inequity in Access to Urban Parks: A Literature Review.” Landscape and Urban Planning 153: 160–169.

Schwarz, Kirsten, Christopher Boone, Sabrina M. Cleland, et al. 2018. “Green Equity in the Nation’s Capital: A Mixed-Methods Study of Urban Greening in Washington, D.C.” Urban Forestry & Urban Greening 32: 1–10.


